---
title: "2024.08_test_chromConverter"
format: html
editor: visual
toc: TRUE
code-fold: TRUE
embed-resources: TRUE
---

## Introduction

The objective of this notebook is to evaluate if the new version of chromConverter produced by Ethan Bass can extract the chromatograms as .txt from .lcd files produced by the [HPLC at Genoscope-CEA](https://github.com/actolonen/Analysis_Lab/tree/main/HPLC).

This notebooks was inspired by [this discussion with Ethan Bass](%5Bhttps://github.com/ethanbass/chromConverter/issues/29) in which he explains that the photodiode array (PDA) stream is empty in the .lcd files produced by our HPLC. This is because the our HPLC lacks a PDA. The two detectors on our HPLC are are refractive index RID-10A (Detector B) and UV/VIS SPD-20A (Detector A).

### Methods

```{r r setup, echo=FALSE, message=FALSE}

library(knitr);
knitr::opts_chunk$set(warning = F, message = F);
knitr::opts_chunk$set(engine.path = list(perl = "/usr/bin/perl"));

library(chromConverter);
library(stringr);
library(tidyverse);
library(readr);
library(gridExtra);

mytheme = theme(axis.text.x = element_text(size = 6), 
                axis.text.y = element_text(size = 6), 
               axis.title.x = element_text(size = 8), 
               axis.title.y = element_text(size = 8),
               strip.text.x = element_text(size = 6),
               legend.position = "bottom", 
               aspect.ratio =0.25);

```

On 06 aug 2024, Ethan updated the master branch of chromConverter to read 2D chromatograms, like those produced by our HPLC. I thus downloaded the updated chromConverter from github:

```         
# install.packages("remotes")
# remotes::install_github("https://github.com/ethanbass/chromConverter/")
```

Run chromConverter function read_shimadzu_lcd() to convert .lcd files to .txt files

```{r convert .lcd file to .txt}

# get a list of .lcd files in the working directory and convert them to data.frames, then print 
# the data.frmes as .txt files.

input.dir = "/home/tolonen/Github/actolonen/Public/Analysis_Lab/HPLC/ChromConverter/Files_LabSolutions";
files.lcd = list.files(path = input.dir, pattern = "lcd");
output.dir = "/home/tolonen/Github/actolonen/Public/Analysis_Lab/HPLC/ChromConverter/Files_chromConverter/";

for (file in files.lcd) 
{
  inputfile = paste(input.dir, file, sep = "/");
  inputfile.base = str_match(string=file, pattern="^[\\d_b]+");
  outputfile = paste(output.dir, inputfile.base, "_chromConverter.txt", sep = "");
  data = read_shimadzu_lcd(
     path = inputfile, format_out = "data.frame", data_format = "long", read_metadata = TRUE);
  write.csv(data, file = outputfile);
}

```

Extract chromatrograms for detector A (UV) and detector B (RID) from LabSolutions .txt files (perl code)

```{perl format programs, eval =F, engine.path = "/usr/bin/perl"}

# perl code to extract detector A and detector B chromatograms.

$dirpath = "/home/tolonen/Github/actolonen/Public/Analysis_Lab/HPLC/ChromConverter/Files_LabSolutions";

# get list of .txt files
opendir(DIR, $dirpath) or die "cannot open directory $dirpath\n";
@allfiles = grep(/\.txt$/, readdir(DIR));

# remove cleaned files from previous runs
foreach $file (@allfiles)
{
 if ($file =~ /detector/)
 {
  $inputfile = "$dirpath"."/"."$file";
  unlink($inputfile);
 }
}

# extract detector B (RID) chromatogram
LOOP:foreach $file (@allfiles)
{
 $inputfilepath = "$dirpath"."/"."$file";
 $file =~ /(.*).txt/;
 $inputfilebase = $1;
 $outputfile = "$inputfilebase"."_detectorB.txt";
 $outputfilepath = "$dirpath"."/"."$outputfile";
 open(IN, "<$inputfilepath") or die "cant open IN\n";
 open(OUT, ">$outputfilepath") or die "cant open OUT\n";
 while ($line = <IN>)
 {
  chomp($line);
  $line =~ s/\r//g;
  unless ($line =~ /LC Chromatogram\(Detector B-Ch1\)/)
  {
   next;
  }
  until ($line =~ /^R.Time/)
  {
    $line = <IN>;
    chomp($line);
    $line =~ s/\r//g;
  }
  print OUT "$line\n";
  $line = <IN>;
  until($line =~ /[a-z]/)
 {
  chomp($line);
  $line =~ s/\r//g;
  print OUT "$line\n";
  $line = <IN>;
 }
next LOOP;
}
close IN;
close OUT;
}

```

Compare chromatograms produced by chromConverter() to those produced by LabSolutions (Shimadzu software) using headerless files.

```{r Compare Chromatograms}

# foreach headerless .txt file in LabSolutions directory, grab file base name, open filehandles to LabSolutions and chromConverter files base file name, plot both files on same plot.

# directories of input files
input.dir.ls = "/home/tolonen/Github/actolonen/Public/Analysis_Lab/HPLC/ChromConverter/Files_LabSolutions";
input.dir.cc = "/home/tolonen/Github/actolonen/Public/Analysis_Lab/HPLC/ChromConverter/Files_chromConverter";

# get list of files in chromConverter directory
files.cc = list.files(path = input.dir.cc, pattern = "txt");

for (file in files.cc) 
{
  print(file);
  path.cc = paste(input.dir.cc, file, sep = "/");
  inputfile.cc.base = str_match(string=file, pattern="^[\\d_b]+");
  file.ls = paste(inputfile.cc.base, "detectorB.txt", sep = "");
  path.ls = paste(input.dir.ls, file.ls, sep = "/");
  data.ls = read.csv(file = path.ls, header=TRUE, sep = "\t");
  data.cc = read.csv(file = path.cc, header=TRUE);

# cleanup input data
data.cc = data.cc %>%
  select(rt, int) %>%
  rename(R.Time = rt, Intensity = int) %>%
  mutate(Format = "ChromConverter");

data.ls = data.ls %>%
  rename(R.Time = R.Time..min.) %>%
  mutate(Format = "LabSolutions") %>%
  mutate(R.Time = as.numeric(R.Time)) %>%
  mutate(Intensity = as.numeric(Intensity)) %>%
  drop_na();

# scale data
data.ratio = max(data.ls$Intensity) / max(data.cc$Intensity);
data.cc = data.cc %>%
  mutate(Intensity = data.ratio * Intensity);

data.all = rbind(data.cc, data.ls);

title.ls = paste("Sample ", inputfile.cc.base, " LabSolutions", sep="");
title.cc = paste("Sample ", inputfile.cc.base, " chromConverter", sep="");

# plot comparison
myplot.ls = ggplot(data.ls, aes(x=R.Time, y=Intensity, group = Format, color = Format)) +
  ggtitle(title.ls)+
  geom_point(size=0.5, color = "red") +
  xlab("Time") + 
  ylab("Intensity") +
  theme_classic() +
  coord_cartesian(xlim=c(0, max(data.ls$R.Time)), ylim=c(0, max(data.ls$Intensity)))+
  scale_x_continuous(breaks=seq(0, max(data.ls$R.Time), max(data.ls$R.Time/10)))+
  scale_y_continuous(breaks=seq(0, max(data.ls$Intensity), max(data.ls$Intensity/8)))+
  mytheme;

myplot.cc = ggplot(data.cc, aes(x=R.Time, y=Intensity, group = Format, color = Format)) +
  ggtitle(title.cc)+
  geom_point(size=0.5, color = "blue") +
  xlab("Time") + 
  ylab("Intensity") +
  coord_cartesian(xlim=c(0, max(data.cc$R.Time)), ylim=c(0, max(data.cc$Intensity)))+
  scale_x_continuous(breaks=seq(0, max(data.cc$R.Time), max(data.cc$R.Time/10)))+
  scale_y_continuous(breaks=seq(0, max(data.cc$Intensity), max(data.cc$Intensity/8)))+
  theme_classic() +
  mytheme;

bothplots = grid.arrange(myplot.ls, myplot.cc, ncol = 1);
bothplots
}

```
