---
title: "Bioscreen: C.phy growth plots"
author: "Fermlab"
format: html
editor: visual
toc: TRUE
code-fold: TRUE
embed-resources: TRUE
---

# Introduction

This notebook analyzes the Bioscreen expt of *C. phytofermentans* WT growing in different media.

# Methods

## Set up

```{r setup, echo=FALSE, message=FALSE}

# General setup code

library(knitr);
knitr::opts_chunk$set(warning = F, message = F);
knitr::opts_chunk$set(engine.path = list(perl = "/usr/bin/perl"));

library(tidyverse);
library(readxl);
library(readr);
library(gridExtra);

mytheme = theme(axis.text.x = element_text(size = 6), 
                axis.text.y = element_text(size = 6), 
               axis.title.x = element_text(size = 8), 
               axis.title.y = element_text(size = 8),
               strip.text.x = element_text(size = 4),
               #legend.position = "bottom", 
               legend.text = element_text(size=4),
               aspect.ratio =0.5);
```

## File I/O

```{r File IO}

# input file
file.in = "growthData_Bioscreen.xlsx";
path.in = "/home/tolonen/Github/actolonen/Public/Analysis_Lab/Growth/Data";
data.in = paste(path.in, file.in, sep = "/");

```

## Organize data for plotting

```{r Organize Bioscreen data for plots}

# output is data.frame = [Hours, Well, OD600, Treatment]

plate.map = read_excel(data.in, sheet = "Information", col_names = TRUE, skip = 3);
growth.data = read_excel(data.in, sheet = "Raw", col_names = TRUE, skip = 1);

# plate map variables
first.well = "12";
last.well = "189";

# parse plate.map
plate.map = plate.map %>%
  filter(!Treatment == "empty") %>%
  mutate(Well = as.character(Well));
  
# calculate elapsed time in hours
time1 = hms("00:00:00"); # start time
growth.data = growth.data %>%
  mutate(Time_hms = hms(Time)) %>%
  mutate(Time_Elapsed = Time_hms - time1) %>%
  mutate(Hours = hour(Time_Elapsed) + 
         (minute(Time_Elapsed) / 60) + 
         (second(Time_Elapsed)/3600));
  
growth.data.long = growth.data %>%
  select(-Blank) %>%
  dplyr::select(Hours, first.well:last.well) %>%
  pivot_longer(cols = first.well:last.well, names_to = "Well", values_to = "OD600");
growth.data.long$OD600 = as.numeric(growth.data.long$OD600);

# remove contaminated wells
#growth.data.long = growth.data.long %>%
#  filter(!Well %in% c("A04", "A05", "A06", "C03", "E02", "G01", "G04", "G05"));

# add treatment info
growth.all = inner_join(growth.data.long, plate.map, by = "Well");


```

# Results

## Plot Individual wells

Plot blanks

```{r Plot blanks, fig.cap = "Fig 1. OD600 of blank treatments (no cells). Each curve is a well."}

growth.blanks = growth.all %>%
  filter(grepl("Blank", Treatment));

plot.blanks = ggplot(growth.blanks, aes(x=Hours, y=OD600, group=Well)) +
  geom_point(size=1, color = "green") +
  geom_line(size=0.1, color='black') +
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.blanks$Hours)), 
    ylim = c(0, max(growth.blanks$OD600) + 0.2))+
  facet_wrap (~Treatment, ncol=2)+
  scale_x_continuous(breaks=seq(0, max(growth.blanks$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.blanks$OD600 + 0.1), 0.05))+
  theme_classic()+
  mytheme;

grid.arrange(plot.blanks, ncol=1);



```

Subtract blanks from wells with cells.

```{r subtract blanks}

growth.cells = growth.all %>%
  filter(grepl("Inhibitor", Treatment));

# subtract initial reading from OD measurements
temp = growth.cells %>%
  slice(1);
initial.time = temp$Hours;

initial.OD = growth.cells %>%
  filter(Hours == initial.time) %>%
  rename(OD600_t0 = OD600) %>%
  select(Well, OD600_t0);
growth.cells = left_join(growth.cells, initial.OD, by = "Well");
growth.cells = growth.cells %>%
  mutate(OD600 = OD600 - OD600_t0) %>%
  select(-OD600_t0);

```

Plot growth in well with cells

```{r plot Growth, fig.cap = "Fig 2. Growth (OD600) of treatments with C.phy. Each curve is a well."}

plot.cells = ggplot(growth.cells, aes(x=Hours, y=OD600, group=Well)) +
  geom_point(size=.05, color = "blue") +
  geom_line(size=0.1, color='black') +
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  facet_wrap (~Treatment, ncol=2)+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 24))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.4))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells, ncol=1);


```

## Plot treatment means +/SD

```{r plot means, fig.cap = "Fig 3. Growth (OD600) of treatments with C.phy. Plot shows treatment means +/-SD."}

growth.cells = growth.cells %>%
  group_by(Hours, Treatment) %>%
  mutate(OD600_mean = mean(OD600)) %>%
  mutate(OD600_sd = sd(OD600)) %>%
  ungroup();
  
plot.cells.means = ggplot(growth.cells, aes(
    x=Hours, 
    y=OD600_mean, 
    ymin = OD600_mean - OD600_sd,
    ymax = OD600_mean + OD600_sd)) +
  geom_point(size=.05, color = "blue") +
  geom_line(size=0.1, color='black') +
  geom_errorbar(width = 0, color = "blue")+
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  facet_wrap (~Treatment, ncol=2)+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 24))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.4))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells.means, ncol=1);

```

## Compare treatment means

C.phy growth at different inhibitor concentrations (1/10 initial dilution)

```{r growth inhibitor 1/10, fig.cap="Fig 4. Growth (OD600) of treatments with C.phy growing in media with different inhibitor concentrations. Plot shows treatment means +/-SD."}

growth.cells.inhibitor = growth.cells %>%
  separate(Treatment, c("Dilution", "Treatment"), sep="-") %>%
  filter(grepl("1/10", Dilution));

plot.cells.inhibitor.10 = ggplot(growth.cells.inhibitor, aes(
    x=Hours, 
    y=OD600_mean, 
    group=Treatment,
    color=Treatment,
    ymin = OD600_mean - OD600_sd,
    ymax = OD600_mean + OD600_sd)) +
  geom_point(size=.05) +
  geom_line(size=0.1, color='black') +
  geom_errorbar(width = 0)+
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.1))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells.inhibitor.10, ncol=1);



```

C.phy growth at different inhibitor concentrations (1/5 initial dilution)

```{r growth inhibitor 1/5, fig.cap="Fig 5. Growth (OD600) of treatments with C.phy growing in media with different inhibitor concentrations. Plot shows treatment means +/-SD."}

growth.cells.inhibitor = growth.cells %>%
  separate(Treatment, c("Dilution", "Treatment"), sep="-") %>%
  filter(grepl("1/5", Dilution));

plot.cells.inhibitor.5 = ggplot(growth.cells.inhibitor, aes(
    x=Hours, 
    y=OD600_mean, 
    group=Treatment,
    color=Treatment,
    ymin = OD600_mean - OD600_sd,
    ymax = OD600_mean + OD600_sd)) +
  geom_point(size=.05) +
  geom_line(size=0.1, color='black') +
  geom_errorbar(width = 0)+
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.1))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells.inhibitor.5, ncol=1);

```

# Conclusions
