---
title: "C.phy growth in different media"
author: "Fermlab"
format: html
editor: visual
toc: TRUE
code-fold: TRUE
embed-resources: TRUE
---

# Introduction

This notebook analyzes the Clario expt of comparing growth of *C. phytofermentans* WT different media.

# Methods

## Set up

```{r setup, echo=FALSE, message=FALSE}

# General setup code

library(knitr);
knitr::opts_chunk$set(warning = F, message = F);
knitr::opts_chunk$set(engine.path = list(perl = "/usr/bin/perl"));

library(tidyverse);
library(readxl);
library(readr);
library(gridExtra);

mytheme = theme(axis.text.x = element_text(size = 6), 
                axis.text.y = element_text(size = 6), 
               axis.title.x = element_text(size = 8), 
               axis.title.y = element_text(size = 8),
               strip.text.x = element_text(size = 4),
               #legend.position = "bottom", 
               legend.text = element_text(size=4),
               aspect.ratio =0.5);
```

## File I/O

```{r File IO}

# input file
file.in = "Croissance_Cphy_M6_clario_27092024.xlsx";
path.in = "/home/tolonen/GenoscopeDrive/Lab_Projects/BMC_ADH/Growth";
data.in = paste(path.in, file.in, sep = "/");

```

## Organize data for plotting

```{r Organize data for plots}

plate.map = read_excel(data.in, sheet = "PLAN PLAQUE", col_names = TRUE, skip = 24);
growth.data = read_excel(data.in, sheet = "RAW DATA", col_names = TRUE, skip = 1);

# plate map variables
first.well = "B02";
last.well = "G09";

# parse plate.map
plate.map = plate.map %>%
  filter(!Treatment == "empty");
  
# parse growth.data
growth.data.long = growth.data %>%
  rename("Seconds" = "...1") %>%
  slice(-1) %>% # delete first row
  #mutate(Time = if_else(grepl("min", Time), true = Time, false =  paste(Time, "0 min", sep = " "))) %>%  # add minutes field where lacking
  #separate(Time, c("hours", "hours.label", "min", "min.label"), sep = " ") %>%
  mutate(Seconds = as.numeric(Seconds)) %>%
  mutate(Hours = Seconds/3600) %>%
  dplyr::select(Hours, first.well:last.well) %>%
  pivot_longer(cols = first.well:last.well, names_to = "Well", values_to = "OD600");
growth.data.long$OD600 = as.numeric(growth.data.long$OD600);

# remove contaminated wells
#growth.data.long = growth.data.long %>%
#  filter(!Well %in% c("A04", "A05", "A06", "C03", "E02", "G01", "G04", "G05"));

# add treatment info
growth.all = inner_join(growth.data.long, plate.map, by = "Well");


```

# Results

## Plot Individual wells

Plot blanks

```{r Plot blanks, fig.cap = "Fig 1. OD600 of blank treatments (no cells). Each curve is a well."}

growth.blanks = growth.all %>%
  filter(!grepl("C.phy", Treatment));

plot.blanks = ggplot(growth.blanks, aes(x=Hours, y=OD600)) +
  geom_point(size=1, color = "green") +
  geom_line(size=0.1, color='black') +
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.blanks$Hours)), 
    ylim = c(0, max(growth.blanks$OD600) + 0.2))+
  facet_wrap (~Treatment, ncol=2)+
  scale_x_continuous(breaks=seq(0, max(growth.blanks$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.blanks$OD600 + 0.1), 0.05))+
  theme_classic()+
  mytheme;

grid.arrange(plot.blanks, ncol=1);



```

Subtract blanks from wells with cells.

```{r subtract blanks}

growth.cells = growth.all %>%
  filter(grepl("C.phy", Treatment));

# subtract initial reading from OD measurements
initial.OD = growth.cells %>%
  filter(Hours == 0) %>%
  rename(OD600_t0 = OD600) %>%
  select(Well, OD600_t0);
growth.cells = left_join(growth.cells, initial.OD, by = "Well");
growth.cells = growth.cells %>%
  mutate(OD600 = OD600 - OD600_t0) %>%
  select(-OD600_t0);

```

Plot growth in well with cells

```{r plot Growth, fig.cap = "Fig 2. Growth (OD600) of treatments with C.phy. Each curve is a well."}

plot.cells = ggplot(growth.cells, aes(x=Hours, y=OD600, group=Well)) +
  geom_point(size=.05, color = "blue") +
  geom_line(size=0.1, color='black') +
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  facet_wrap (~Treatment, ncol=2)+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.1))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells, ncol=1);


```

## Plot treatment means +/SD

```{r plot means, sd, fig.cap = "Fig 3. Growth (OD600) of treatments with C.phy. Plot shows treatment means +/-SD."}

growth.cells = growth.cells %>%
  group_by(Hours, Treatment) %>%
  mutate(OD600_mean = mean(OD600)) %>%
  mutate(OD600_sd = sd(OD600)) %>%
  ungroup();
  
plot.cells.means = ggplot(growth.cells, aes(
    x=Hours, 
    y=OD600_mean, 
    ymin = OD600_mean - OD600_sd,
    ymax = OD600_mean + OD600_sd)) +
  geom_point(size=.05, color = "blue") +
  geom_line(size=0.1, color='black') +
  geom_errorbar(width = 0, color = "blue")+
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  facet_wrap (~Treatment, ncol=2)+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.1))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells.means, ncol=1);

```

## Compare treatment means

C.phy utilization of choline, ethanolamine as carbon sources

```{r Carbon sources, fig.cap="Fig 4. Growth (OD600) of treatments with C.phy growing in M6 medium with different carbon sources. Plot shows treatment means."}

growth.cells.carbon = growth.cells %>%
  filter(grepl("M6_1|M6_2|M6_3|M6_4", Treatment));

plot.cells.carbon = ggplot(growth.cells.carbon, aes(
    x=Hours, 
    y=OD600_mean, 
    group=Treatment,
    color=Treatment))+
    #ymin = OD600_mean - OD600_sd,
    #ymax = OD600_mean + OD600_sd)) +
  geom_point(size=.05) +
  geom_line(size=0.1, color='black') +
  #geom_errorbar(width = 0, color = "blue")+
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.1))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells.carbon, ncol=1);



```

C.phy utilization of choline, ethanolamine as nitrogen sources

```{r Nitrogen sources, fig.cap="Fig 4. Growth (OD600) of treatments with C.phy growing in M6 medium with different nitrogen sources. Plot shows treatment means."}

growth.cells.nitrogen = growth.cells %>%
  filter(grepl("M6_1|M6_5|M6_6|M6_7|M6_8", Treatment));

plot.cells.nitrogen = ggplot(growth.cells.nitrogen, aes(
    x=Hours, 
    y=OD600_mean, 
    group=Treatment,
    color=Treatment))+
    #ymin = OD600_mean - OD600_sd,
    #ymax = OD600_mean + OD600_sd)) +
  geom_point(size=.05) +
  geom_line(size=0.1, color='black') +
  #geom_errorbar(width = 0, color = "blue")+
  theme_bw() +
  xlab("Hours") + 
  ylab("OD(600)") +
  coord_cartesian(
    xlim = c(0, max(growth.cells$Hours)), 
    ylim = c(-0.05,  max(growth.cells$OD600) + 0.1))+
  scale_x_continuous(breaks=seq(0, max(growth.cells$Hours), 12))+
  scale_y_continuous(breaks=seq(0, max(growth.cells$OD600 + 0.1), 0.1))+
  theme_classic()+
  mytheme;

grid.arrange(plot.cells.nitrogen, ncol=1);
```

# Conclusions
